---
title: "CBF_Logistic_Regression"
author: "Krish Shah"
date: "2023-11-15"
output: html_document
---

Clear the environment and load in all required packages.

```{r}
rm(list=ls())
suppressPackageStartupMessages({
  library(dplyr)
  library(sjPlot)
  library(tidyr)
  library(pROC)
})
```

Read in TAP REDcap export.

```{r}
df <- read.csv("/Users/krishshah/Downloads/TAPElectronicDataCap-TAPPullASLWHR_DATA_2023-10-15_1301.csv")
```

Pre-processing TAP data. Make sure to add usability variable when it is available.

```{r}
df$session_id <- if_else(df$session_id=="",NA,df$session_id)
merged_df <- df %>%
  group_by(vmac_id, redcap_event_name) %>%
  summarise_all(~ ifelse(all(is.na(.)), NA, na.omit(.)[1]))
df <- merged_df
colnames(df) <- gsub("_", ".", colnames(df))

df$diagnosis.factor <- ifelse(df$diagnosis == 1, "Normal",
                              ifelse(df$diagnosis == 2, "MCI",
                                     ifelse(df$diagnosis == 3, "Dementia",
                                            ifelse(df$diagnosis == 4, "Ambiguous At Risk", NA))))

df$epoch <- ifelse(df$redcap.event.name == "enrollmentbaseline_arm_1", 1,
                              ifelse(df$redcap.event.name == "1year_followup_arm_1", 2,
                                     ifelse(df$redcap.event.name == "2year_followup_arm_1",
                                            3,df$redcap.event.name)))


df <- dplyr::filter(df,!is.na(asl.rest.grey.matter.hct))
df <- df %>% 
  dplyr::group_by(vmac.id) %>%
  dplyr::arrange(epoch,.by_group = TRUE) %>% 
  dplyr::filter(row_number()==1)


table(df$diagnosis.factor)

df$diagnosis.factor <- factor(df$diagnosis.factor,
                              levels = c("Normal", "Ambiguous At Risk", "MCI", "Dementia"),
                              ordered = F)


df$diagnosis.factor.ordered <- factor(df$diagnosis.factor,
                              levels = c("Normal", "Ambiguous At Risk", "MCI", "Dementia"),
                              ordered = T)

df$apoe <- ifelse(grepl("E4", df$alleles), "Yes", "No")

df$apoe4pos.factor <- factor(df$apoe,
                              levels = c("No", "Yes"),
                              ordered = F)

df$cdr.factor <- factor(df$cdrglob,
                        levels = c(0,0.5,1,2,3),
                        ordered = F)

df$education <- df$educ
df$sex.factor <- if_else(df$sex==1,"Male","Female")
df$sex.factor <- as.factor(df$sex.factor)
```

Load in latest VMAP merge.

```{r}
df_vmap <- readRDS("/Users/krishshah/Downloads/MAP_fm_f20230906_m20230906.rds")
```

Pre-processing VMAP data and selecting first observation.

```{r}
df_vmap <- dplyr::filter(df_vmap,(epoch>=3 | map.id>=337))
df_vmap <- filter(df_vmap, asl.rest.usable.hct.factor == "Yes")
df_vmap <- df_vmap %>% 
  dplyr::group_by(vmac.id) %>%
  dplyr::arrange(epoch,.by_group = TRUE) %>% 
  dplyr::filter(row_number()==1)
```

Merging VMAP and TAP data.

```{r}
df_tst <- plyr::rbind.fill(df,df_vmap)
table(df_tst$cdr.factor)
table(df_tst$diagnosis.factor)
table(df_tst$apoe4pos.factor)
```

Selecting all variables of interest from merged dataframe.

```{r}
df_sel <- dplyr::select(df_tst,vmac.id,epoch,cdr.factor,age,sex.factor,education,asl.rest.pcu.precuneus.hct,
                        diagnosis.factor, apoe4pos.factor, ma.total.intracranial.vol, raceethnicity.factor,
                        asl.rest.putamen.hct, asl.rest.prg.precentral.gyrus.hct, asl.rest.hippocampus.hct,
                        asl.rest.itg.inferior.temporal.gyrus.hct,
                        asl.rest.po.parietal.operculum.hct, asl.rest.morg.medial.orbital.gyrus.hct,
                        asl.rest.mfg.middle.frontal.gyrus.hct, asl.rest.grey.matter.hct.scov,
                        np.cvlt.ldcr, np.cvlt.ldfr, np.inhibit, asl.rest.grey.matter.hct)
```

Dropping samples that are missing a variable.

```{r}
selected_columns <- c("cdr.factor", "age", "education", "asl.rest.pcu.precuneus.hct", "apoe4pos.factor", 
                      "sex.factor", "asl.rest.putamen.hct", "asl.rest.grey.matter.hct")
df_sel <- df_sel %>% drop_na(any_of(selected_columns))
```

Binarize cdr.factor.
```{r}
table(df_sel$cdr.factor)

df_sel$cdr.factor.binarized <- droplevels(df_sel$cdr.factor)
df_sel$cdr.factor.binarized <- if_else(df_sel$cdr.factor.binarized==0.5 | df_sel$cdr.factor.binarized==2 | df_sel$cdr.factor.binarized==3,factor(1, levels=levels(df_sel$cdr.factor.binarized)),df_sel$cdr.factor.binarized)
df_sel$cdr.factor.binarized <- droplevels(df_sel$cdr.factor.binarized)

table(df_sel$cdr.factor.binarized)
table(df_sel$cdr.factor.binarized, df_sel$apoe4pos.factor)
```

Test logistic regression model and ROC curve.
```{r}
df_sel$asl.rest.grey.matter.hct.resid <- residuals(lm(asl.rest.grey.matter.hct ~ asl.rest.putamen.hct, data = df_sel))
logmodel <- glm(cdr.factor.binarized ~ asl.rest.grey.matter.hct.resid + age + sex.factor + education + apoe4pos.factor, data=df_sel, family="binomial")
summary(logmodel)
roc(df_sel$cdr.factor.binarized, logmodel$fitted.values,
    plot=TRUE, legacy.axes=TRUE, percent=TRUE, print.auc=TRUE, print.auc.y=40, col="red")
```