---
title: "CBF and Cognition"
author: "Krish Shah"
date: "2023-10-21"
output: html_document
---
Clear the environment and load in required packages.
```{r}
rm(list=ls())
require(tidyr)
require(dplyr)
require(nlme)
require(sjPlot)
```

## Filtering
Read in latest VMAP merge.
```{r}
df <- readRDS("/Users/krishshah/Downloads/MAP_fm_f20230906_m20230906.rds")
```

Create variable categories.
```{r}
idVars <- c("map.id", "epoch")
timeVar <- c("np.date.time")            
usabilityVarsBase <- c("asl.rest.usable.hct")
usabilityVarsLong <- c("asl.rest.usable.hct")
covariates <- c("apoe4pos.factor","age",
                 "raceethnicity.factor","fsrp.minus.age.points","sex.factor",
                 "education","diagnosis.factor","ma.total.intracranial.vol", "cdr.factor")
predictors <- c("asl.rest.grey.matter.hct", "asl.rest.pcu.precuneus.hct", "asl.rest.prg.precentral.gyrus.hct", "asl.rest.hippocampus.hct",
                "asl.rest.itg.inferior.temporal.gyrus.hct",
                "asl.rest.po.parietal.operculum.hct", "asl.rest.morg.medial.orbital.gyrus.hct",
                "asl.rest.mfg.middle.frontal.gyrus.hct", "asl.rest.putamen.hct")
outcomes <- c("np.bnt","np.anim","np.executive.composite","np.memory.composite")
colsLongit <- c(idVars,timeVar,usabilityVarsLong,outcomes,"NumDays")
allVars <- c(idVars,timeVar,usabilityVarsBase,usabilityVarsLong,covariates,predictors,outcomes)
```

Collect all of the variables of interest.
```{r}
df <- dplyr::select(df,all_of(allVars))
```

Calculate # of days since baseline.
```{r}
df$scan.date.days <- as.Date(df[,timeVar])
startdate <- as.Date("01/01/2000","%d/%m/%Y")
df$NumDays  <- as.numeric(difftime(df$scan.date.days,startdate ,units="days"))
```

Remove NA and create baseline dataframe.
```{r}
df <- filter(df, df$diagnosis.factor != "Dementia")
df_base <- tidyr::drop_na(df)
```

Remove unusable vars and obtain first observation for df_base + resid variable for each predictor.
```{r}
for (usabilityVar in usabilityVarsBase){
  print(usabilityVar)
  df_base <- df_base[df_base[usabilityVar]==1,]
}

df_base<-df_base%>%
  dplyr::group_by(map.id) %>%
  dplyr::arrange(epoch,.by_group = TRUE) %>% 
  dplyr::filter(row_number()==1)

for (pred in predictors) {
  pred_base <- paste(pred, ".resid", sep = "")
  formula <- as.formula(paste(pred, "~ I(asl.rest.putamen.hct)", sep = ""))
  df_base[[pred_base]] <- residuals(lm(formula, data = df_base, na.action = na.exclude))
}
```

Create longitudinal dataframe, drop NAs, and remove unusable vars.
```{r}
df_longit <- dplyr::select(df,all_of(colsLongit))
df_longit <- tidyr::drop_na(df_longit)

for (usabilityVar in usabilityVarsLong){
  print(usabilityVar)
  df_longit <- df_longit[df_longit[usabilityVar]==1,]
}
```

Confirm mapIDs from both df_longit and df_base are matching.
```{r}
df_longit <- df_longit %>% filter(map.id %in% df_base$map.id)
```

Remove first observation from df_longit.
```{r}
df_longit <- dplyr::anti_join(df_longit,df_base,by=c("map.id","epoch"))
```

Create final (merged) dataframe.
```{r}
df_merged <- plyr::rbind.fill(df_base,df_longit)
df <- df_merged
```

Get change in days column by subtracting first observation from every subsequent observation.
```{r}
df <- df %>% 
  dplyr::group_by(map.id) %>%
  dplyr::arrange(epoch,.by_group = TRUE) %>% 
  dplyr::mutate(diffDays = NumDays - dplyr::first(NumDays))
```

.base for all covariates and predictors + diffYr variable.
```{r}
predictors_resid <- c("asl.rest.grey.matter.hct.resid", "asl.rest.pcu.precuneus.hct.resid", "asl.rest.prg.precentral.gyrus.hct.resid", "asl.rest.hippocampus.hct.resid",
                "asl.rest.itg.inferior.temporal.gyrus.hct.resid",
                "asl.rest.po.parietal.operculum.hct.resid", "asl.rest.morg.medial.orbital.gyrus.hct.resid",
                "asl.rest.mfg.middle.frontal.gyrus.hct.resid", "asl.rest.putamen.hct.resid")
for (covar in c(covariates,predictors,predictors_resid)){
  covar_base <-paste(covar,".base",sep="")
  df <- df %>% 
    dplyr::group_by(map.id) %>%
    dplyr::arrange(epoch,.by_group = TRUE) %>% 
    dplyr::mutate( !!covar_base := dplyr::first(get(covar)))
}
df$diffYr <- df$diffDays/365.25

```

## Modeling
Linear mixed effects model for longitudinal EF without residualization using precuneus CBF.
```{r}
ctrl <- lmeControl(msMaxIter = 100000, opt="optim")
mdl1 <- lme(np.executive.composite~diagnosis.factor.base+asl.rest.pcu.precuneus.hct.base*diffYr+apoe4pos.factor.base+fsrp.minus.age.points.base+
              age.base+sex.factor.base+education.base+
              raceethnicity.factor.base, random=~1+diffYr | map.id, na.action=na.omit, 
            control=ctrl, data=df)
summary(mdl1)
plot_model(mdl1, type="eff",terms=c("diffYr","asl.rest.pcu.precuneus.hct.base"))
```

Linear mixed effects model for longitudinal EF with residualization using precuneus CBF.
```{r}
ctrl <- lmeControl(msMaxIter = 100000, opt="optim")
mdl1 <- lme(np.executive.composite~diagnosis.factor.base+asl.rest.pcu.precuneus.hct.resid.base*diffYr+apoe4pos.factor.base+fsrp.minus.age.points.base+
              age.base+sex.factor.base+education.base+
              raceethnicity.factor.base, random=~1+diffYr | map.id, na.action=na.omit, 
            control=ctrl, data=df)
summary(mdl1)
plot_model(mdl1, type="eff",terms=c("diffYr","asl.rest.pcu.precuneus.hct.resid.base"))
```

Linear mixed effects model for longitudinal EF with residualization using precuneus CBF. Plot for interaction effects.
```{r}
ctrl <- lmeControl(msMaxIter = 100000, opt="optim")
mdl1 <- lme(np.executive.composite~diffYr*asl.rest.pcu.precuneus.hct.resid.base*diagnosis.factor.base+apoe4pos.factor.base+fsrp.minus.age.points.base+
            age.base+sex.factor.base+education.base+
            raceethnicity.factor.base, random=~1+diffYr | map.id, na.action=na.omit, 
            control=ctrl, data=df)
summary(mdl1)
plot_model(mdl1, type="int")
```

Linear mixed effects model for longitudinal EF with residualization using precuneus CBF. Plot for interaction effects.
```{r}
ctrl <- lmeControl(msMaxIter = 100000, opt="optim")
mdl1 <- lme(np.executive.composite~diffYr*asl.rest.pcu.precuneus.hct.resid.base*apoe4pos.factor.base+fsrp.minus.age.points.base+diagnosis.factor.base+
            age.base+sex.factor.base+education.base+
            raceethnicity.factor.base, random=~1+diffYr | map.id, na.action=na.omit, 
            control=ctrl, data=df)
summary(mdl1)
plot_model(mdl1, type="int")
```

Linear mixed effects model for longitudinal EF with residualization using precuneus CBF. Plot for interaction effects.
```{r}
ctrl <- lmeControl(msMaxIter = 100000, opt="optim")
mdl1 <- lme(np.executive.composite~diffYr*asl.rest.pcu.precuneus.hct.resid.base*apoe4pos.factor.base+fsrp.minus.age.points.base+diagnosis.factor.base+
            age.base+sex.factor.base+education.base+
            raceethnicity.factor.base, random=~1+diffYr | map.id, na.action=na.omit, 
            control=ctrl, data=df)
summary(mdl1)
plot_model(mdl1, terms=c("diffYr","asl.rest.pcu.precuneus.hct.resid.base","apoe4pos.factor.base"), type = "eff")
```

Linear mixed effects model for longitudinal memory with residualization using precuneus CBF. Plot for interaction effects.
```{r}
ctrl <- lmeControl(msMaxIter = 100000, opt="optim")
mdl1 <- lme(np.memory.composite~diffYr*asl.rest.pcu.precuneus.hct.resid.base*age.base+apoe4pos.factor.base+fsrp.minus.age.points.base+diagnosis.factor.base+sex.factor.base+education.base+
            raceethnicity.factor.base, random=~1+diffYr | map.id, na.action=na.omit, 
            control=ctrl, data=df)
summary(mdl1)
plot_model(mdl1, terms = c("diffYr", "asl.rest.pcu.precuneus.hct.resid.base", "age.base"),
           type = "eff")
```

Linear mixed effects model for longitudinal EF with and without residualization using precuneus CBF.
```{r}
ctrl <- lmeControl(msMaxIter = 100000, opt="optim")
mdl1 <- lme(np.executive.composite~diagnosis.factor.base+asl.rest.pcu.precuneus.hct.resid.base*diffYr+apoe4pos.factor.base+fsrp.minus.age.points.base+
              age.base+sex.factor.base+education.base+
              raceethnicity.factor.base, random=~1+diffYr | map.id, na.action=na.omit, 
            control=ctrl, data=df)
mdl2 <- lme(np.executive.composite~diagnosis.factor.base+asl.rest.pcu.precuneus.hct.base*diffYr+apoe4pos.factor.base+fsrp.minus.age.points.base+
              age.base+sex.factor.base+education.base+
              raceethnicity.factor.base, random=~1+diffYr | map.id, na.action=na.omit, 
            control=ctrl, data=df)
mdl3 <- lme(np.executive.composite~diagnosis.factor.base+(asl.rest.pcu.precuneus.hct.base/asl.rest.putamen.hct.base)*diffYr+apoe4pos.factor.base+fsrp.minus.age.points.base+
              age.base+sex.factor.base+education.base+
              raceethnicity.factor.base, random=~1+diffYr | map.id, na.action=na.omit, 
            control=ctrl, data=df)
anova(mdl3,mdl2, mdl1)
```













